Concurrent Programming with CXL

### 2020-05-21 book/paper.tex
Starting to work on reader/writer

\documentclass{report}
\usepackage{fullpage}
\usepackage{graphicx}
\usepackage{framed}
\usepackage{tcolorbox}
\usepackage{verbatim}
\usepackage{comment}

@@

Here \texttt{@ncs} and \texttt{@cs} are \emph{labels}, each identifying
a location in the CXL machine code.  The first thing we need to
ensure is that there cannot be two processes at the critical section.
We would like to place an assertion at the \texttt{@cs} label that
specifies that only the current process can be there.  CXL in fact
supports this.  It has a method \texttt{atLabel $L$}, where $L$
is the atom of the label (in this case, \texttt{.cs}).  The
method returns a bag of name tags of processes executing at that
label.  The bag is represented by a dictionary that maps each element
in the bag to the number of times the element appears in the bag.

@@

\chapter{Semaphores}

>>>>

\documentclass{report}
\usepackage{fullpage}
\usepackage{graphicx}
\usepackage{framed}
\usepackage{tcolorbox}
\usepackage{verbatim}
\usepackage{comment}
\usepackage{amsmath}

@@

Here \texttt{@ncs} and \texttt{@cs} are \emph{labels}, each identifying
a location in the CXL machine code.  The first thing we need to
ensure is that there cannot be two processes at the critical section.
We would like to place an assertion at the \texttt{@cs} label that
specifies that only the current process can be there.  CXL in fact
supports this.  It has a method \texttt{atLabel $L$}, where $L$
is the atom of the label (in this case, \texttt{.cs}).  The
method returns a bag of name tags of processes executing at that
label.  The bag is represented by a dictionary that maps each element
in the bag to the number of times the element appears in the bag.
Method \texttt{atLabel} only exists for specification purposes---do not
use it in normal code.

@@

\chapter{Reader/Writer Locks}

\begin{figure}
\begin{code}
\verbatiminput{../RWbusy.cxl}
\end{code}
\caption{Busy-Waiting Reader/Writer Lock implementation.}
\label{fig:rwbusy}
\end{figure}

Locks are useful when accessing a shared data structure.  By preventing
more than one process from accessing the data structure at the same
time, conflicting access are avoided.  However, not all concurrent
accesses conflict, and opportunities for concurrency may be lost,
hurting performance.  One important case is when multiple processes
are simply reading the data structure.  There is no conflict, and
all reads should be allowed to proceed concurrently.  In many
applications, reads are the majority of all accesses.

What we want is a special kind of lock that allows either one writer
or one or more readers to be in the critical section.  This is called
a \emph{reader/writer lock}.
We will explore various ways of implementing reader/writer locks in
this chapter and future ones.

Figure~\ref{fig:rwbusy} presents a solution that uses a
single (ordinary) lock and two counters: one that maintains the number
of readers and one that maintains the number of writers.
The program shows two kinds of processes: \texttt{reader()} and
\texttt{writer()}.  The critical section is spread between two
labels: readers access \texttt{@rcs} and writers access \texttt{@wcs}.
The specification is that if a reader is at label \texttt{@rcs},
no writer is allowed to be at label \texttt{@wcs}.  Vice versa, if
a writer is at label \texttt{@wcs}, no reader is allowed to be at
label \texttt{@rcs} \emph{and} there cannot be another writer at
label \texttt{@wcs}.

A reader process that wants to enter \texttt{@rcs} first waits
until there are no writers: \texttt{nwriters == 0}.
If so, it increments the number of readers.  Similarly, a writer
process waits until there are no readers \emph{or} writers.
If so, it increments the number of writers.
The important invariants in this code are:
\begin{itemize}
\item $n$ readers at \texttt{@rcs} $\Rightarrow \mathtt{nreaders} = n$,
\item $n$ writers at \texttt{@rcs} $\Rightarrow \mathtt{nwriters} = n$,
\item $(\mathtt{nreaders} \ge 0 \land \mathtt{nwriters} = 0) \lor
    (\mathtt{nreaders} = 0 \land 0 \le \mathtt{nwriters} \le 1)$.
\end{itemize}
It is easy to see that the invariants hold and imply the specification.
